---
title: "Package Tutorial for psbpHMM"
output: rmarkdown::html_vignette
date: August 17, 2021
author: Lauren Hoskovec
vignette: >
  %\VignetteIndexEntry{Package Tutorial for psbpHMM}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The R package psbpHMM provides several functions for fitting covariate-dependent infinite hidden Markov models (iHMM) via the probit stick-breaking process (PSBP). For more details on this method, see [Hoskovec et al.]. The following functions are available in the package psbpHMM: 

|function          |         brief description of usage |
|------------------|------------------------------------|
simData | simulate data to reproduce simulation in [Hoskovec et al.] | 
miHMM | fit homogeneous infinite hidden Markov model to one or more time series | 
mciHMM | fit covariate-dependent infinite hidden Markov model to one or more time series |
getJointSimResults | summarize simulation output from jointly fit models |
getIndepSimResults | summarize simulation output from independently fit models |
|fitDPMM | fit Dirichlet process mixture model to one or more time series |



In this vignette, we show how to install the package, simulate data, fit three versions of our proposed approach, and summarize the MCMC output. The package contains everything needed to conduct a simulation study similar to that in [Hoskovec et al.]. 

# Install the Package #

First, we need to load the following package dependencies. Use install.packages() to install them, if needed.   

```{r, message = FALSE}
library(Rcpp) 
library(RcppArmadillo)  
library(parallel)
library(gdata)
library(invgamma)
library(gtools)
library(mvtnorm)
library(matrixcalc)
library(tmvmixnorm)
library(rlist)
library(truncnorm)
library(mvnfast)
```

The package can then be installed from gitHub. 

```{r, message = FALSE, results = FALSE}
devtools::install_github("lvhoskovec/psbpHMM", build_vignettes = TRUE)
library(psbpHMM)
```

# Simulate Data #

The package contains functions to simulate data similar to the simulation study in [Hoskovec et al.], using the `simData` function. By default, `simData` generates 3-dimensional multivariate time series for 20 subjects, where each subject has 288 time points, 6 true clusters, and shared temporal patterns. 

The `simData` function takes in the following parameters: `n` is the number of time series, `t.max` is the length of each time series, `p` is the dimension of the data, and `K` is the true number of hidden states shared among all time series. The parameter `trend` allows us to specify shared or distinct temporal trends among the time series, and the parameter `missingLevel` allows us to specify the proportion of missing data, which is split evenly between missing at random (MAR) and below the limit of detection (LOD). Based on these parameters, data are simulated as described in Section 3 in [Hoskovec et al.]. 

Here we simulate a data set with 5$\%$ missing observations. 

```{r, cache=TRUE}
set.seed(42412)
dat = simData(n=20, t.max = 288, K = 6, p = 3, trend = "shared", missingLevel = 0.05)
```

In these data, 2.5$\%$ of the data are MAR and 2.5$\%$ of the data are below the LOD. Hence, the fixed and known LOD is set at the 0.025 quantile of the pooled data set. 

The object `dat` is a list with 6 elements: `y` is a list where each element contains an n by p matrix of simulated data for each subject with MAR observations denoted by NA and below LOD observations denoted by -Inf; `y.complete` is a list of the complete data for each subject before missing observations were removed; `z.true` is a list of the true hidden state trajectories for each subject; `K` is the true number of hidden states among all time series; `mu.true` is a K by p matrix of the true state-specific means; and `lod` is a list of the limits of detection for each subject. 

We can plot the data to see what it looks like. We plot three exposures for one time series. 

```{r, cache = TRUE, fig.width=8, fig.height=5}
par(mfrow = c(1,3))
plot(1:288, dat$y[[1]][,1], col = dat$z.true[[1]], pch = 19, ylab = "exposure 1", 
     xlab = "time", type = "l", xaxt = "n")
axis(side=1, at=seq(0,288+24,24), labels = TRUE)
plot(1:288, dat$y[[1]][,2], col = dat$z.true[[1]], pch = 19, ylab = "exposure 2", 
     xlab = "time", type = "l", xaxt = "n")
axis(side=1, at=seq(0,288+24,24), labels = TRUE)
plot(1:288, dat$y[[1]][,3], col = dat$z.true[[1]], pch = 19, ylab = "exposure 3", 
     xlab = "time", type = "l", xaxt = "n")
axis(side=1, at=seq(0,288+24,24), labels = TRUE)
```

# Models #

## Statistical Approach ##

The package psbpHMM provides functionality to fit three types of iHMMs. Briefly, let $i=1, \ldots, n$ denote subjects and $t = 1, \ldots, T$ denote time points, so that $\mathbf{y}_{it}$ is observed data for subject $i$ at time $t$ and $z_{it}$ is the hidden state for subject $i$ at time $t$.

All models have the same emission distribution for the data conditional on the hidden states. The emission distribution is 

$$
\begin{eqnarray}
f(\textbf{y}_{it}|z_{it}=k) & \equiv & \text{N}(\boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k)  \\
\boldsymbol{\mu}_k|\boldsymbol{\Sigma}_k & \sim & \text{N}\left( \boldsymbol{\mu}_0, \frac{1}{\lambda}\boldsymbol{\Sigma}_k \right) \\
\boldsymbol{\Sigma}_k & \sim & \text{Inverse Wishart} (\nu, \mathbf{R}), 
\end{eqnarray}
$$

where $\boldsymbol{\mu}_0$, $\lambda$, $\nu$, and $\mathbf{R}$ are fixed hyperparameters. 

The difference among the three models lies in the specification of the the transition distribution. 

First (1), is a model without any covariates in the transition distribution; that is, the probability of transitioning from state j to state k is: 


$$
\begin{eqnarray}
P(z_{it} = k | z_{i,t-1} = j) = \Phi (\alpha_{jk} ) \prod_{l < k} \lbrace 1 - \Phi (\alpha_{jl})\rbrace. 
\end{eqnarray}
$$

Second (2) is a model with covariate-dependent transitions. The covariate-dependent transition distribution is 

$$
\begin{eqnarray}
P(z_{it} = k | z_{i,t-1} = j, \mathbf{x}_{it}) &=& \Phi (\alpha_{jk} + \mathbf{x}_{it}^T\boldsymbol{\beta}_k ) \prod_{l < k} \lbrace 1 - \Phi (\alpha_{jl} +  \mathbf{x}_{it}^T\boldsymbol{\beta}_l )\rbrace,
\end{eqnarray}
$$

where $\mathbf{x}_{it}$ are covariates for subject $i$ at time $t$. 

Third (3) is a model with subject-specific effects of the covariates in the transition distribution to account for repeated sampling days for the same individual (3). Here, the transition distribution is 

$$
\begin{eqnarray}
P(z_{i_st} = k | z_{i_s,t-1} = j, \mathbf{x}_{i_st}) = \Phi(\alpha_{jk} + \mathbf{x}_{i_st}^T\boldsymbol{\beta}_k + \mathbf{x}_{i_st}^T\boldsymbol{\gamma}_{ik}  )\prod_{l < k} \lbrace 1 - \Phi(\alpha_{jl} + \mathbf{x}_{i_st}^T\boldsymbol{\beta}_l + \mathbf{x}_{i_st}^T\boldsymbol{\gamma}_{il} )\rbrace,
\end{eqnarray}
$$

where $i_s$ denotes sampling day $s$ for subject $i$. 

The model is completed with the following hyperpriors 

$$
\begin{eqnarray}
\alpha_{jk}|\sigma^2_\alpha & \sim & \text{N}(\mu_\alpha, \sigma^2_{\alpha}) \text{ for } j \neq k \\
\sigma^{-2}_\alpha & \sim & \text{Gamma}(a_1, b_1) \\
\alpha_{jj}|m_\alpha, v_\alpha & \sim & \text{N}(m_\alpha, v_{\alpha}) \\
m_\alpha & \sim & \text{N}(m_0, v_0) \\
v_\alpha^{-1} & \sim &  \text{Gamma}(a_2, b_2) \\
\boldsymbol{\beta}_k & \sim & \text{N}(\boldsymbol{\mu}_\beta, \boldsymbol{\Sigma}_\beta) \\
\boldsymbol{\gamma}_{ik} & \sim & \text{N}(\boldsymbol{\mu}_\gamma, \kappa^2 \boldsymbol{\Sigma}_\gamma) \\
\kappa^{-2} & \sim & \text{Gamma}(\alpha_\kappa,\beta_\kappa).
\end{eqnarray}
$$

## Function Details ##

Model (1) is fit using the function `miHMM` and models (2) and (3) are fit using the function `mciHMM`. The parameters in `miHMM` are a subset of those in `mciHMM`, so we describe how to use the function `mciHMM` here. 

The function `mciHMM` takes many parameters. We first describe the most common ones. 

To begin, `niter` is the total number of MCMC iterations and `nburn` is the number of burn-in iterations. `y` is an $n$-element list, where $n$ is the number of subjects. `y` represents the emission data and may contain missing observations denoted by NA (for MAR data) or -Inf (for data below LOD). Each element of `y` is a $T \times p$ matrix, representing data for each subject, where $T$ is the length of each time series and $p$ is the dimension of the data. `X` is an $n$-element list of time-varying covariates for each subject. Each element of `X` is a $T \times q$ matrix where $q$ is the dimension of the covariate data. The ordering of subjects in `X` and `y` must be the same. `rmlist` is a vector of integer labels for the repeated sampling days. Each integer corresponds a subject and an element of `y` and `X`; unique integers represent unique subjects and repeated integers represent repeated sampling days for the same subject. `missing` is logical and specifies whether `y` contains missing observations or not. If missing is true, `lod` is an $n$-element list of lower limits of detection for each of the $p$ exposures and `len.imp` is the number of imputations to save. Imputations will be save at equally spaced iterations post burn-in. `K.start` is the starting number of hidden states; the default is 12. `priors` is the list of prior hyperparameters. The prior options for this model include the following: 

|control parameter | model parameter | default value | description      |
|------------------|-----------------|---------------|------------------|
|mu0         |$\boldsymbol{\mu}_0$  | 0             |mean parameter for Normal prior on $\boldsymbol{\mu}_k$ |
|lambda| $\lambda$ | 0 | scale parameter for Normal prior on $\boldsymbol{\mu}_k$  | 
|nu|$\nu$ |  1 |  degrees of freedom for Inverse Wishart prior on $\boldsymbol{\Sigma}_k$  |
|R | $\mathbf{R} | $\mathbf{I}_p$ | matrix parameter for Inverse Wishart prior on $\boldsymbol{\Sigma}_k$ |
|mu.alpha |$\mu_\alpha$   | 0       |mean parameter for Normal priors on $\alpha_{jk$, $j \neq k$|
|a1| $a_1$ | 1 |  shape parameter for gamma prior on $\sigma_\alpha^{-2}$ |
|b1|$b_1$  |  1 |   rate parameter for gamma prior on $\phi_2^{-2}$ |
|m0 |$m_0$ |0| mean parameter for Normal prior on $m_\alpha$ |
|v0 |$v_0$ |1| variance parameter for Normal prior on $m_\alpha$|
|a2|$a_2$ |  1 | shape parameter for gamma prior on $v_\alpha^{-1}$ |
|b2|$b_2$ |  1 |  rate parameter for gamma prior on $v_\alpha^{-1}$  |
|mu.beta | $\boldsymbol{\mu}_\beta$ | $\mathbf{0}$ | mean parameter for Normal prior on $\boldsymbol{\beta}_k$ | 
|Sigma.beta| $\boldsymbol{\Sigma}_\beta$ | $\mathbf{I}$ | variance parameter for Normal prior on $\boldsymbol{\beta}_k$ | 
|mu.gamma | $\boldsymbol{\mu}_\gamma$ | $\mathbf{0}$ | mean parameter for Normal prior on $\boldsymbol{\gamma}_{ik}$ | 
|Sigma.gamma| $\boldsymbol{\Sigma}_\gamma$ | $\mathbf{I}$ | variance parameter for Normal prior on $\boldsymbol{\beta}_k$ | 
|a.kappa | $\alpha_\kappa$ | 1 | shape parameter for gamma prior on $\kappa^{-2}$ 
|b.kappa | $\alpha_\kappa$ | 1 | rate parameter for gamma prior on $\kappa^{-2}$ 

Next we describe the parameters for tuning the Metropolis-Hastings independence sampler when there are missing observations. Details can be found in the Supporting Information of [Hoskovec et al.]. The independence MH sampler uses the following the parameters 

|control parameter | model parameter | default value | description      |
|------------------|-----------------|---------------|------------------|
|tau2         |$\tau^2$  | 0.25             |variance parameter for Normal proposal on lower triangular elements |
|a.tune| $a_\delta$ | 10 | shape parameter for inverse gamma proposal on diagonal elements  | 
|b.tune|$b_\delta$ |  1 |  rate parameter for inverse gamma proposal on diagonal elements  |

There is an option to use a resolvent kernel to update the parameters of the lower triangular matrix in the decomposition of $\boldsymbol{\Sigma}_k$. The parameter `resK` is a logical parameter; if set to TRUE then the resolvent kernel will be used and `eta.star` specifies the parameter of the geometric distribution for the resolvent kernel. The default value for `eta.star` is 5. 

Finally, we include some parameters used to conduct a simulation study and evaluation of hidden state estimation and imputation. `z.true` is an $n$-element list of true hidden state trajectories for each subject. `mu.true` is a $K_{\text{true}} \times p$ matrix of true state-specific means, where $K_{\text{true}}$ is the true number of hidden states. `ycomplete` is a list of the same dimensions as `y` that contains the true values for the missing observations in `y`. 

The parameter `holdout` is rarely used. It is the same class and dimension of `y` but specifies which observations are in the holdout data set for evaluation imputation on an incomplete data set. In `holdout`, 0 denotes the data is observed, 1 denotes MAR, and 2 denotes below LOD. The only reason to specify `holdout` is when evaluating the multiple imputation approach on an incomplete data set; that is, when `ycomplete` still contains missing observations. `holdout` specifies the additional data that was removed for validation, but not the observations that are missing in `ycomplete`. 

# Examples #

We demonstrate each model here. First we fit the homogenous PSPB iHMM to the simulated data. Since our data has missing observations, we specify the tuning parameters for the MH independence sampler. Here we run the model for 100 iterations with 50 burn-in and save 25 imputations post burn-in. On your own data, be sure to run the model for a long enough chain to reach convergence. To save time, we have provided the results of the model fits in the package, which can be loaded with the data() function. We show the code for fitting the model below, and then show how to load the results. 


```{r, eval = FALSE, cache = TRUE}
fit_no_cov = miHMM(niter = 100, nburn = 50, y = dat$y, ycomplete = dat$y.complete, 
      z.true = dat$z.true, lod = dat$lod, mu.true = dat$mu.true, missing = TRUE, 
      tau2 = 0.25, a.tune = 10, b.tune = 1, resK = TRUE, eta.star = 5, 
      len.imp = 25)
```


```{r}
data("fit_no_cov")
```


First, let's look at the MH acceptance rates. We want to set the tuning parameters so that the MH acceptances rates are between 0.1 and 0.4. 

```{r, cache = TRUE}
fit_no_cov$MH.arate; fit_no_cov$MH.lamrate
```

Next, we fit the covariate-dependent PSBP. To incorporate covariates, we simulate cyclical temporal trends via a harmonic function of time. We will assume the time points for each time series all correspond to the same time of day, though they may take place on different days. Hence, the harmonic function of time will be identical for all time series. The current version of this model only permits time-varying covariates. 

```{r, cache = TRUE}
t.max = 288
n = 20
transT = seq(1:t.max)/t.max*2*pi
X1 = cbind(sin(transT), cos(transT), sin(2*transT), cos(2*transT))
X = list()
for(i in 1:n){
  X[[i]] = X1
}
q = ncol(X[[1]])
```

Now that we have covariates for each subject, we can fit the model. 

```{r, eval = FALSE, cache = TRUE}
fit_cyclical = mciHMM(niter = 100, nburn = 50, y = dat$y, ycomplete = dat$y.complete, 
                      X = X, z.true = dat$z.true, lod = dat$lod, mu.true = dat$mu.true, 
                      missing = TRUE, tau2 = 0.25, a.tune = 10, b.tune = 1, resK = TRUE, eta.star = 5, 
                      len.imp = 25)
```


```{r}
data("fit_cyclical")
```

Last, we fit the model with subject-specific effects of covariates. To do so, we specify the parameter `rmlist` to provide categorical indicators of the repeated time series for each subject. Each integer corresponds to a unique subject. The order of the categorical variables in the vector `rmlist` must be the same as the order of the time series in the lists `X` and `y` data. The following `rmlist` represents 4 unique subjects each with 5 repeated sampling days. 

```{r}
rmlist = c(1,1,1,1,1,2,2,2,2,2,3,3,3,3,3,4,4,4,4,4)
```

```{r, eval = FALSE, cache = TRUE}
fit_ss = mciHMM(niter = 100, nburn = 50, y = dat$y, rmlist = rmlist, ycomplete = dat$y.complete, 
             X = X, z.true = dat$z.true, lod = dat$lod, mu.true = dat$mu.true, 
             missing = TRUE, tau2 = 0.25, a.tune = 10, b.tune = 1, resK = TRUE, eta.star = 5, 
             len.imp = 25)
```

```{r}
data("fit_ss")
```

# Analyze Results #

The package provides the function `getJointSimResults` to return the evaluation criteria measures used in our simulation study in [Hoskovec et al.]. This function returns the hamming distance (hamming), which is a measure of the distance between the estimated hidden state trajectories and the true hidden state trajectories, and the MSE for state-specific means (mu.mse) averaged over the iterations post burn-in. It also returns the MSE for MAR data (mar.mse) and data below the LOD (lod.mse) average over the specified number of imputations post burn-in. The best performing method will have the smallest hamming distance, MSE for state-specific means, and MSE for imputations. Here we show the summary for each of the models we fit. 

Model (1): no covariates

```{r, cache=TRUE}
getJointSimResults(fit_no_cov)
```

Model (2): cyclical function of time as covariates

```{r, cache = TRUE}
getJointSimResults(fit_cyclical)
```

Model (3): subject-specific cyclical function of time 

```{r, cache = TRUE}
getJointSimResults(fit_ss)
```


From this abbreviated simulation study demonstration, the subject-specific cyclical model best estimated the hidden state trajectories and state-specific means with lowest Hamming distance and MSE for $\mu$. The subject-specific cyclical model also had the lowest MSE for MAR imputations, closely followed by the cyclical model without random effects. The cyclical model without random effects had the lowest MSE for below LOD imputations, followed by the subject-specific model. The model without any covariates performed worst in all measures. 


We provide some functions for post-processing the results. The following code applies to all three of our models. We will show on `fit_cyclical` only. First, we can calculate the most optimal hidden state trajectory using the draws-based latent structure optimization method described by Dahl (2006). We provide a wrapper function for the `dlso` function from the `salso` R package. 

```{r, cache = TRUE}
zbest = dlso_wrapper(fit_cyclical)
```

Given the most optimal hidden state trajectories, we can calculate model-averaged estimates of the state-specific means with `modelAvemu`. This function must take in the exposure data as a matrix so we turn the list into a matrix here. 

```{r, cache = TRUE}
ymatrix=NULL
for(i in 1:n){
  ymatrix = rbind(ymatrix, dat$y[[i]])
}
mubest = modelAveMu(fit = fit_cyclical, zbest1=zbest, ymatrix = ymatrix)
```

# Additional Models #

In our simulation study, we compare our proposed joint approach to the models fit independently to each time series. Fit the cyclical model independently to each time series by

```{r, eval = FALSE, cache = TRUE}
fit_indep_cyclical = mclapply(1:20, FUN = function(i){
  mciHMM(niter = 100, nburn = 50, y = dat$y[[i]], ycomplete = dat$y.complete[[i]], 
                      X = list(X[[i]]), z.true = dat$z.true[[i]], lod = list(dat$lod[[i]]), 
                      mu.true = dat$mu.true, missing = TRUE, 
                      tau2 = 0.25, a.tune = 10, b.tune = 1, resK = TRUE, eta.star = 5, 
                      len.imp = 25)
})
```

```{r}
data("fit_indep_cyclical")
```

Similalrly, fit the model with no covaraites independently to each time series by 


```{r, eval = FALSE, cache = TRUE}
fit_indep_nocov = mclapply(1:20, FUN = function(i){
  miHMM(niter = 100, nburn = 50, y = dat$y[[i]], ycomplete = dat$y.complete[[i]], 
                      z.true = dat$z.true[[i]], lod = list(dat$lod[[i]]), 
                      mu.true = dat$mu.true, missing = TRUE, 
                      tau2 = 0.25, a.tune = 10, b.tune = 1, resK = TRUE, eta.star = 5, 
                      len.imp = 25)
})
```


```{r}
data("fit_indep_nocov")
```

To summarize the results from the independently fit models, use the function `getIndepSimResults`. This function averages the hamming distance and MSE for state-specific means and imputations over each of the `n` time series. Here, you must also include the emission data as a parameter. 

```{r, cache = TRUE}
getIndepSimRsults(fit1 = fit_indep_cyclical, y = dat$y)
```

```{r, cache = TRUE}
getIndepSimRsults(fit1 = fit_indep_nocov, y = dat$y)
```

Finally, we include in our package a function to fit a truncated Dirichlet process mixture model (DPMM) jointly to multiple time series. Here, `K.start` specifies the maximum number of hidden states allowed. We set `K.start = 50`. 

```{r, eval = FALSE, cache = TRUE}
fit_dpmm = fitDPMM(niter = 100, nburn = 50, y = dat$y, ycomplete = dat$y.complete, 
             K.start = 50, z.true = dat$z.true, lod = dat$lod, 
             mu.true = dat$mu.true, missing = TRUE, 
             tau2 = 0.25, a.tune = 10, b.tune = 1, resK = TRUE, eta.star = 3, len.imp = 25)
```

```{r}
data("fit_dpmm")
```

Get the results from the joint DPMM using `getJointSimResults`.

```{r, cache = TRUE}
getJointSimResults(fit_dpmm)
```

## Contact Information ##

For questions or comments, please contact the package maintainer Lauren Hoskovec at lvheck@rams.colostate.edu. 


